---
title: "Analyze single-cell data"
author: "Allen Goodman and Shantanu Singh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyze single-cell data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to analyze single-cell data from a
morphological profiling experiment. 

The images were analyzed using [CellProfiler](http://cellprofiler.org).

This vignette assumes that the single-cell data `110000106828_sub_1.sqlite` have 
been created using `ingest` and stored in `~/Downloads`.


```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(magrittr)

```


# Load data

First, load the data! The data is contained in 4 tables named `Image`, 
`Cytoplasm`, `Cells`, and `Nuclei`. The code below joins these tables to 
create a single table named `object`. 


```{r load, message=FALSE}
backend <- file.path(Sys.getenv("HOME"), "Downloads", "110000106828_sub_1.sqlite")

db <- src_sqlite(path = backend)

image <- tbl(src = db, "image")

object <-
  tbl(src = db, "Cells") %>%
  inner_join(tbl(src = db, "Cytoplasm"),
                    by = c("TableNumber", "ImageNumber", "ObjectNumber")) %>%
  inner_join(tbl(src = db, "Nuclei"),
                    by = c("TableNumber", "ImageNumber", "ObjectNumber"))

object %<>% inner_join(image, by = c("TableNumber", "ImageNumber"))

```

In this table, the measurement columns start with `Nuclei_`, `Cells_`, or 
`Cytoplasm_`. 

```{r}
variables <-
  colnames(object) %>%
  stringr::str_subset("^Nuclei_|^Cells_|^Cytoplasm_")
```

How many variables?

```{r}
print(length(variables))
```

How many cells?

```{r}
object %>% 
  count() %>%
  knitr::kable(caption = "No. of cells")
```

Before aggregating the data, let's filter it down to one well and plot the 
histogram of a single feature:

```{r fig.width=5, fig.height=5}
object %>% 
  filter(Image_Metadata_Well == "A04") %>%
  select(Nuclei_Intensity_IntegratedIntensity_Hoechst) %>% 
  collect() %>% {
    ggplot(., aes(Nuclei_Intensity_IntegratedIntensity_Hoechst)) + 
      geom_histogram(binwidth = .1)
    }

```

# Summarize measurements per well

Now let's summarize the data by grouping by well and computing averages. 

```{r message=FALSE}
profiles <-
  cytominer::aggregate(
    population = object,
    variables = variables,
    strata = c("Image_Metadata_Barcode", "Image_Metadata_Well"),
    operation = "mean"
  )

profiles %<>%
  collect()

```

How many wells?

```{r}
profiles %>% 
  count() %>%
  knitr::kable(caption = "No. of wells")
```

Let's plot the relationship between a pair of variables from this summarized data

```{r fig.width=6, fig.height=5, message=FALSE}
p <- 
  ggplot(profiles, aes(Cells_AreaShape_Area, Nuclei_AreaShape_Area)) + 
  geom_point()

print(p)
```

